---
title: "HW 01"
author: "Jelani Butler"
format: 
  html:
    embed-resources: true
toc: true
---

## 0 - Setup

```{r setup}
if (!require("pacman")) 
  install.packages("pacman")

# use this line for installing/loading
pacman::p_load(tidyverse,accidents,nyc_marathon,county)
pacman::p_load()

devtools::install_github("tidyverse/dsbox")

ggplot2::theme_set(ggplot2::theme_minimal(base_size = 14))

options(width = 65)


knitr::opts_chunk$set(
  fig.width = 7,        # 7" width
  fig.asp = 0.618,      # the golden ratio
  fig.retina = 3,       # dpi multiplier for displaying HTML output on retina
  fig.align = "center", # center align figures
  dpi = 300             # higher dpi, sharper image
)
pacman::p_load(here)
pacman::p_load(dplyr)
pacman::p_load(lubridate)
pacman::p_load(openintro)
pacman::p_load(scales)

```


## 1 - Road traffic accidents in Edinburgh

```{r label-me-1}

```
#I wanted to see the data set before I got started
accidents <- read_csv(here("accidents.csv"))
glimpse(accidents)

#now create the weekend dummy variables for the weekend and weekday


accidents <- accidents|>
  mutate(
    Type_day_cat = case_when(
      day_of_week %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday") ~ "Weekday",
      day_of_week %in% c("Saturday", "Sunday") ~ "Weekend",
    
    )
  
)

#I want to look at the new variable 

glimpse(accidents)

```
now I have to extract the hour from the time coloumn section. looking at stack overflow "https://stackoverflow.com/questions/77692894/how-to-extract-hour-from-datetime-column-in-r" I was able to find the lubridate library has the extraction tools necessary to get the hour. I also have to add it to the table as "hour".

```

time <- hms::as_hms(accidents$time)
accidents <- accidents|>
  mutate(
      hour = hour(hms::as_hms(accidents$time)))

glimpse(accidents)

# ok time to create the new graph. looking at the output it is clear that we need to use the geom_density I'm not too sure if you wanted the exact colors as the ones in the picture.

ggplot(accidents, aes(x = hour, fill = severity, color = severity))+
  geom_density(alpha = 0.6, position = "identity")+
  facet_wrap(~ Type_day_cat, ncol=1)+
  labs(
  title = "Number of accidents throughout the day 
  By day of week and severity",
  x = "Time of Day",
  y= "Density"
)

# interpretation: there seems to be alot of slight and serious accidents around 6 pm on the weekday (which makes sense because that's around the time of rush hour.) what's even more interesting is that the fatal accidents seem to be more prevalent during the early mornings.
#the weekend seems to have similar affects for slight and serious accidents but there seems to be very few fatal accidents on the weekends.
  

## 2 - NYC marathon winners

```{r label-me-2}

```

data(nyc_marathon)

glimpse(nyc_marathon)


sum(is.na(nyc_marathon))

```
I can tell that I'm going to have to supress the warnings so I'll do that now
```

suppressWarnings(data) 

```
make the charts
```
suppressWarnings(data) 

ggplot(nyc_marathon, aes(x = time)) +
  geom_histogram(bins = 30, fill = "skyblue", color = "black") +
  labs(title = "The Winning Times of Marathon Runners", x = "Time", y = "Number")




ggplot(nyc_marathon, aes(y = time_hrs)) +
  geom_boxplot(fill = "blue") +
  labs(title = "Box Plot of The Winning Times of Marathon Runners", y = "Hours")

```
YOu can see more of a binned group of runner's time in the histogram chart. You also see the distribution of the runner's winning times in the histogram chart.  

YOu can see the quartiles and the winning times for each runner in the box plot. You can also see the outliers and how far they are away from the 25th, 50th, and 75th percentile of winning runner's times.
```

ggplot(nyc_marathon, aes(x= division, y = time_hrs, fill=division)) +
  geom_boxplot()+
  scale_fill_manual(values = c("Men"= "Blue", "Women"="Pink"))
  labs(title = "Box Plot of The Winning Times of Marathon Runners", y = "Hours")

```
Based on the box plots it seems that the winning men overall had faster winning times than the women. There are less outliers for the men (7) than the women (8) for the slower runners.    
```

```
C.)

it is clear that the legend here is not needed. So I will remove it. this will improve the data to ink ratio becuse it will have more data ink and less non-data ink.     
```
ggplot(nyc_marathon, aes(x= division, y = time_hrs, fill=division)) +
  geom_boxplot(show.legend = FALSE)+
  scale_fill_manual(values = c("Men"= "Blue", "Women"="Pink"))
  labs(title = "Box Plot of The Winning Times of Marathon Runners", y = "Hours")

```
D.)
I have to make a point plot to make sure that i have the times and years in the chart. I have to change the color from women from pink to red so you will be able to clearly see the observations. what's visible in this plot that's different from the other plots is the exact times and the sepeficic years of the men and women's winning times. You are not able to see this in the other charts.     
```
suppressWarnings(nyc_marathon) 
ggplot(nyc_marathon) +
  geom_point(aes(x= year, y=time, color =division))+
  scale_color_manual(values = c("Men"= "Blue", "Women"="Red"))+
  labs(title = "The Winning Times of Marathon Runners","x= year", y = "time")

## 3 - US counties
```
A.)

Looking at the code, it is clear that you are trying to add layers to the code and for some reason the second layer that you have is another chart with different variables which won't work. SO no it does not make sense and the code will NOT work.   
```
```
B.)

It is easier to compare poverty leverls across people from the first table in my opinion. The biggest advantage is that in this case, seeing the chart horizontally, allows you to see the trends within the dataset. The big point is that you have to be careful on how you place your variables because it can hinder, misinform, or, make it difficult to provide meaningful
observations. It is also important to know where "~" is placed when using the facet_grid feature.  
  
```
```
C.)

```

library(openintro)
data(county)
glimpse(county)

#Plot A
ggplot(county,(aes(x = homeownership, y = poverty))) + 
  geom_point()
   labs(
  title = "Plot A"
)

  
#plot B
ggplot(county,(aes(x = homeownership, y = poverty))) + 
  geom_point()+
  geom_smooth(method = "loess", se = FALSE)+
   labs(
  title = "Plot B"
)

  
#Plot C
ggplot(county,(aes(x = homeownership, y = poverty))) + 
  geom_point()+
  geom_smooth(aes(color=metro),method = "loess", se = FALSE, show.legend = FALSE)+
  labs(
  title = "Plot C"
)

  
# plot D

ggplot(county, aes(x = homeownership, y = poverty)) +
  geom_smooth(aes(color = metro), method = "loess", se = FALSE,show.legend = FALSE)+
  geom_point()+
   labs(
  title = "Plot D"
)


# plot E


ggplot(county, aes(x = homeownership, y = poverty)) +
  geom_point(aes(color = metro)) +
  geom_smooth(aes(color = metro, linetype = metro), method = "loess", se = FALSE) +
  scale_color_manual(values = c("yes" = "blue", "no" = "black", "NA" = "gray")) +
  scale_linetype_manual(values = c("yes" = "dotted", "no" = "solid", "NA" = "solid")) +
  labs(
  title = "Plot E"
)


# Plot F

ggplot(county, aes(x = homeownership, y = poverty)) +
  geom_point(aes(color = metro)) +
  geom_smooth(aes(color = metro, linetype = metro), method = "loess", se = FALSE) +
  scale_linetype_manual(values = c("yes" = "dotted", "no" = "solid", "NA" = "solid")) +
  labs(
  title = "Plot F"
  )

# Plot G

  ggplot(county,(aes(x = homeownership, y = poverty, color = metro))) + 
  geom_point()+
  geom_smooth(method = "loess", se = FALSE, color= "blue")+
   labs(
  title = "Plot G"
)

# Plot H

  ggplot(county, aes(x = homeownership, y = poverty)) +
  geom_point(aes(color = metro)) +
  labs(
  title = "Plot H"
  )
  
## 4 - Rental apartments in SF

```
A.)

Based on the code below you see a relationship between credit card balance, income and marital/student status. 

There is a strong relationship between income and credit card balance in general. 

The more income you have, the more you will be allowed to borrow by an lending compnay. 

This relationship can vary depending on the martial status and whether or not the borrower is in school. 

Normally, if someone chooses to attend school, they are traditionally foregoing current income with the potential to recieve more income in the future with the new degree.

Also it is traditionally assumed that if you have a spouse in todat's modern society, it's assumed that both members of the relationship are working, also called a 2 income household. 

Therefore, furthering the notion that if you are married, you  will have both members of the household working, which increases the income, therefore increases the income and credit limit. 

If you are a traditional student (full time, usually no employment status) you will have less income, therefore a lower credit limit. 

```
credit <- read_csv(here("credit.csv"))



credit <- credit |>
  mutate(
  color_group = ifelse(student == "No", "Top", "Bottom"),
  shape_group = ifelse(student == "No", "circle", "triangle")
  )

glimpse(credit)

custom_labels <- labeller(
  student = c("No" = "Student: No", "Yes" = "Student: Yes"),
  married = c("No" = "Married: No", "Yes" = "Married: Yes")
)


ggplot(credit, aes(x = income, y = balance, color = color_group, shape = shape_group)) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  geom_point()+
  facet_grid(student ~ married, labeller = custom_labels) +
  scale_color_manual(values = c("Top" = "blue", "Bottom" = "red")) +
  scale_shape_manual(values = c("circle" = 16, "triangle" = 17)) +
  scale_x_continuous(labels = label_dollar(scale = 1, suffix = "K")) +
  scale_y_continuous(labels = label_dollar()) +
  labs(
    x = "Income",
    y = "Credit card balance"
  )+
  theme(legend.position = "none")


```
B.) As explained in part A. both married and student are useful predictors in addition to income for predicting credit card balance.

As stated before, if you are married, the notion is in today's society both members of the household are working. Therefore, there will be more income and a higher credit limit.

On the other hand, if there are traditional students, they are most likely not in the labor force currently. therefore, less income, and a lower credit balance. 

If you don't agree with the standars assumptions, then one can believe that it might not be useful predictors. For example, students might not work, but their accounts might be tied to wealthy parents, therefore breaking the assumptions stated in part A. 

Or if a married household is only a single working household, where one of the significant others are for example dealing with the upbrinigng of children. The assumption also breaks. 

```

```
C.) 
code wise, it's the same thing but you have to create some new code for the new var, credit utilization. 

Then you change the format of the credit balance to be a percent and not a dollar value. 

```

credit <- credit |>
  mutate(
  credit_utilization = balance / limit,
  color_group = ifelse(student == "No", "Top", "Bottom"),
  shape_group = ifelse(student == "No", "circle", "triangle")
  )

glimpse(credit)

custom_labels <- labeller(
  student = c("No" = "Student: No", "Yes" = "Student: Yes"),
  married = c("No" = "Married: No", "Yes" = "Married: Yes")
)


ggplot(credit, aes(x = income, y = credit_utilization, color = color_group, shape = shape_group)) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  geom_point()+
  facet_grid(student ~ married, labeller = custom_labels) +
  scale_color_manual(values = c("Top" = "blue", "Bottom" = "red")) +
  scale_shape_manual(values = c("circle" = 16, "triangle" = 17)) +
  scale_x_continuous(labels = label_dollar(scale = 1, suffix = "K")) +
  scale_y_continuous(labels = label_percent(accuracy = 1, suffix = "%")) +
  labs(
    x = "Income",
    y = "Credit card utilization"
  )+
  theme(legend.position = "none")


```
D.) 
Based on what we have in part C there's not alot of differnece, it mainly comes down to the amount of income that you have. the more income you have, the less you will need to use your credit. Therfore the lower the credit utilization. Based on this data, marital status nor being a student had much of an effect on the relationship between income and credit blaance. 
```


## 5 - Napoleon’s march.





```Works cited
jeffreyohene. “How to Extract Hour from Datetime Column in R.” Stack Overflow, 20 Dec. 2023, https://stackoverflow.com/questions/77692894/how-to-extract-hour-from-datetime-column-in-r. Accessed 30 May 2025.

Islands. “Multiple Plots on 1 Page in R.” Stack Overflow, 16 May 2014, https://stackoverflow.com/questions/23692026/multiple-plots-on-1-page-in-r. Accessed 30 May 2025.

sieste. “How to Suppress Warnings Globally in an R Script.” Stack Overflow, 24 Apr. 2013, https://stackoverflow.com/questions/16194212/how-to-suppress-warnings-globally-in-an-r-script. Accessed 30 May 2025.

“R Programming Cheat Sheet.” Dataquest, https://www.dataquest.io/cheat-sheet/r-programming-cheat-sheet/. Accessed 30 May 2025.
```

